---
title: "scCertify: Explainable Confidence Scoring for Single-Cell Annotations"
author: "Jaya Surya Doddetipalli"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{scCertify: Explainable Confidence Scoring for Single-Cell Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)

## Optional dependencies used to build the fully worked example below.
## If any are unavailable, the corresponding chunks are skipped rather
## than failing, but every chunk still shows exactly what would run.
have_seurat  <- requireNamespace("Seurat", quietly = TRUE)
have_ucell   <- requireNamespace("UCell", quietly = TRUE)
have_fnn     <- requireNamespace("FNN", quietly = TRUE)
have_entropy <- requireNamespace("entropy", quietly = TRUE)

run_demo <- have_seurat && have_ucell && have_fnn && have_entropy
```

# Introduction

Single-cell RNA sequencing (scRNA-seq) enables the identification of diverse cellular populations from complex biological samples. Numerous computational methods have been developed to assign cell-type labels to individual cells; however, most annotation methods provide little information regarding the reliability of each prediction.

**scCertify** complements existing annotation workflows by estimating confidence for predicted cell identities rather than performing annotation itself. The package combines multiple independent sources of evidence into a calibrated confidence score together with interpretable explanations.

The confidence framework integrates four complementary components:

- Marker gene enrichment
- Neighborhood agreement
- Entropy-based uncertainty
- Doublet information

The resulting confidence scores can be used to identify uncertain annotations, prioritize manual review, improve downstream analyses, and increase the transparency of automated annotation workflows.

# Installation

Install the development version using `BiocManager`.

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("scCertify")
```

Load the package.

```{r load-package}
library(scCertify)
```

# Workflow Overview

A typical scCertify workflow consists of:

1. Predict cell identities using an external annotation method.
2. Supply a marker gene database.
3. Compute marker consistency scores.
4. Measure neighborhood agreement.
5. Estimate entropy-based uncertainty.
6. Combine evidence into confidence scores.
7. Interpret confidence using explanation utilities.

The package supports both **Seurat** and **SingleCellExperiment** objects through a common interface.

To make every step below runnable, we first build a small simulated
`Seurat` object with predicted labels already assigned, mirroring what
your data would look like after running an external annotation method
(e.g. `SingleR`, `Azimuth`, or manual marker-based annotation).

```{r simulate-data, eval=run_demo}
set.seed(42)

n_genes <- 50
n_cells <- 40

counts <- matrix(
    rpois(n_genes * n_cells, lambda = 5),
    nrow = n_genes,
    ncol = n_cells
)

rownames(counts) <- paste0("Gene", seq_len(n_genes))
colnames(counts) <- paste0("Cell", seq_len(n_cells))

object <- Seurat::CreateSeuratObject(counts = counts)

object$predicted_label <- rep(
    c("T cell", "B cell"),
    each = n_cells / 2
)

object <- Seurat::NormalizeData(object, verbose = FALSE)
object <- Seurat::FindVariableFeatures(object, verbose = FALSE)
object <- Seurat::ScaleData(object, verbose = FALSE)
object <- Seurat::RunPCA(object, npcs = 10, verbose = FALSE)

object
```

# Input Requirements

The input object should already contain predicted cell labels, stored
in a metadata column (default: `predicted_label`).

For Seurat objects:

```{r seurat-metadata, eval=run_demo}
head(object@meta.data)
```

For SingleCellExperiment objects, the same column would live in `colData()`:

```{r sce-metadata, eval=FALSE}
head(SummarizedExperiment::colData(object))
```

A marker gene database is supplied as a named list, with one element per
predicted cell type:

```{r marker-list}
markers <- list(
  "T cell" = c("Gene1", "Gene2", "Gene3"),
  "B cell" = c("Gene4", "Gene5", "Gene6")
)

markers
```

(In the simulated object above, `predicted_label` already matches
these names exactly; `match_labels()`, demonstrated later, is what lets
looser real-world labels such as `"T_cells"` still match `"T cell"`.)

# Marker Consistency Score

Marker consistency evaluates whether a cell expresses genes expected for its predicted identity. scCertify uses the UCell algorithm to compute robust rank-based enrichment scores.

```{r marker-score, eval=run_demo}
marker_scores <- marker_score(
    object,
    markers
)

head(marker_scores)
```

Higher scores indicate stronger molecular support for the assigned annotation.

# Neighborhood Agreement Score

Correctly annotated cells generally occur near cells of the same identity in low-dimensional space.

```{r neighbor-score, eval=run_demo}
neighbor_scores <- neighbor_score(
    object
)

head(neighbor_scores)
```

Neighborhood agreement ranges from 0 to 1, where larger values indicate greater agreement with nearby cells.

# Entropy-Based Uncertainty

Entropy provides a quantitative measure of annotation uncertainty. Unlike `marker_score()` and `neighbor_score()`, `entropy_score()` operates on a **numeric score matrix** (for example, per-cell prediction probabilities across candidate labels from your annotation method) rather than directly on the Seurat/SingleCellExperiment object.

```{r entropy-score, eval=have_entropy}
score_matrix <- matrix(
    runif(20),
    nrow = 5
)

entropy_score(score_matrix)
```

For our simulated dataset, we treat each cell as having a two-class
prediction-probability row (as if from an external classifier), compute
entropy on that matrix, and normalize it to `[0, 1]` before storing it
on the object as `entropy_norm`. This is the column `cell_certify()`
expects to find.

```{r entropy-norm, eval=run_demo}
prob_matrix <- matrix(
    runif(n_cells * 2, min = 0.1, max = 1),
    ncol = 2
)

raw_entropy <- entropy_score(prob_matrix)

object$entropy_norm <- raw_entropy / log(ncol(prob_matrix))

head(object$entropy_norm)
```

Low entropy indicates confident predictions, whereas high entropy suggests ambiguous assignments.

scCertify does not currently ship its own doublet detector; supply a
`doublet_score` column from a tool such as `scDblFinder` if available.
Here we simulate one so `cell_certify()` has something to use:

```{r doublet-score, eval=run_demo}
object$doublet_score <- runif(n_cells, min = 0, max = 0.3)
```

If `entropy_norm` or `doublet_score` is missing entirely, `cell_certify()` will still run: it substitutes zeros and issues a warning rather than failing.

# Computing Confidence Scores

The final confidence score integrates all available evidence via `cell_certify()`, which internally recomputes marker and neighborhood scores, combines them with `entropy_norm` and `doublet_score`, and adds the results back onto the object. You do not need to call `marker_score()` and `neighbor_score()` separately in practice&mdash;they were shown above only to illustrate what each component measures.

The default weighting scheme is:

| Component | Weight |
|:----------|------:|
| Marker enrichment | 35% |
| Neighborhood agreement | 35% |
| Entropy | 20% |
| Doublet penalty | 10% |

```{r cell-certify, eval=run_demo}
object <- cell_certify(
    object,
    markers
)

head(
    object@meta.data[
        ,
        c(
            "predicted_label",
            "marker_score",
            "neighbor_score",
            "entropy_norm",
            "doublet_score",
            "confidence_score",
            "confidence_class"
        )
    ]
)
```

The resulting metadata include:

- `marker_score`
- `neighbor_score`
- `confidence_score`
- `confidence_class`

`classify_confidence()` is what turns raw scores into these classes, and can be used on its own with any numeric vector of scores:

```{r confidence-classification}
scores <- c(0.15, 0.55, 0.92)

data.frame(
    score = scores,
    class = classify_confidence(scores)
)
```

For SingleCellExperiment objects, the equivalent metadata would be added to `colData()`:

```{r sce-metadata-result, eval=FALSE}
head(
    SummarizedExperiment::colData(object)
)
```

# Custom Confidence Thresholds

Confidence classes can be customized by changing the moderate/high thresholds passed to `cell_certify()`:

```{r custom-thresholds, eval=run_demo}
object_custom <- cell_certify(
    object,
    markers,
    moderate_threshold = 0.4,
    high_threshold = 0.75
)

table(object_custom$confidence_class)
```

# Explaining Confidence

Two small helper functions support the scoring pipeline and can be used independently:

```{r helper-functions}
calibrate_confidence(
  c(-1, 0, 1)
)

match_labels(
  "T_cells",
  c("T_cell", "B_cell")
)
```

`calibrate_confidence()` converts raw scores into percentile ranks, and `match_labels()` matches loosely-formatted predicted labels (e.g. underscores, plural forms) against the names of your marker database.

To understand the major contributors to a single cell's confidence score, use `explain_confidence()`:

```{r explain-confidence, eval=run_demo}
explain_confidence(
    object,
    cell_id = "Cell1"
)
```

For a fuller, printed summary of a single cell&mdash;including its predicted label and every component score&mdash;use `explain_cell()`:

```{r explain-cell, eval=run_demo}
explain_cell(
    object,
    cell_id = "Cell1"
)
```

Possible explanations include:

- Weak marker enrichment
- Neighborhood disagreement
- High entropy
- Possible doublet contamination

# Interpreting Results

High-confidence annotations generally exhibit:

- Strong marker enrichment
- Strong neighborhood agreement
- Low entropy
- Low doublet scores

Low-confidence annotations frequently correspond to:

- Ambiguous transcriptional states
- Rare cell populations
- Doublets
- Poor-quality cells
- Incorrect annotations

These cells may benefit from manual inspection before downstream analyses.

# Compatibility

scCertify supports both:

- Seurat
- SingleCellExperiment

allowing integration into CRAN- and Bioconductor-based workflows.

# Summary

scCertify provides an explainable confidence scoring framework for single-cell annotation. By combining marker enrichment, neighborhood agreement, entropy, and doublet information, it quantifies annotation reliability while providing interpretable explanations for each confidence assignment.

The package is intended to complement existing annotation pipelines and assist with quality control, annotation refinement, downstream filtering, and biological interpretation.

# References

- Andreatta M, Carmona SJ (2021). *UCell: Robust and scalable single-cell gene signature scoring*. Computational and Structural Biotechnology Journal.

- Aran D, Looney AP, Liu L, et al. (2019). *Reference-based analysis of lung single-cell sequencing reveals a transitional profibrotic macrophage*. Nature Immunology.

# Session Information

```{r session-info}
sessionInfo()
```
