---
title: "scCertify: Explainable Confidence Scoring for Single-Cell Annotations"
author: "Jaya Surya Doddetipalli"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{scCertify: Explainable Confidence Scoring for Single-Cell Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)

have_sce <- requireNamespace("SingleCellExperiment", quietly = TRUE)
have_sce_tools <- requireNamespace("SummarizedExperiment", quietly = TRUE)
have_ucell <- requireNamespace("UCell", quietly = TRUE)
have_fnn <- requireNamespace("FNN", quietly = TRUE)
have_entropy <- requireNamespace("entropy", quietly = TRUE)

run_demo <- have_sce && have_sce_tools && have_ucell &&
    have_fnn && have_entropy
```

# Introduction

Single-cell RNA sequencing (scRNA-seq) enables the identification of diverse cellular populations from complex biological samples. Numerous computational methods have been developed to assign cell-type labels to individual cells; however, most annotation methods provide little information regarding the reliability of each prediction.

**scCertify** complements existing annotation workflows by estimating confidence for predicted cell identities rather than performing annotation itself. The package combines multiple independent sources of evidence into a calibrated confidence score together with interpretable explanations.

The confidence framework integrates four complementary components:

- Marker gene enrichment
- Neighborhood agreement
- Entropy-based uncertainty
- Doublet information

The resulting confidence scores can be used to identify uncertain annotations, prioritize manual review, improve downstream analyses, and increase the transparency of automated annotation workflows.

# Installation

Install the development version using `BiocManager`.

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("scCertify")
```

Load the package.

```{r load-package}
library(scCertify)
```

# Workflow Overview

A typical scCertify workflow consists of:

1. Obtain cell-type predictions from an external annotation method.
2. Store the predicted labels in a `SingleCellExperiment` object.
3. Supply a marker gene database.
4. Compute marker consistency scores.
5. Provide a low-dimensional representation for neighborhood scoring.
6. Estimate entropy-based uncertainty.
7. Combine the evidence into confidence scores.
8. Interpret confidence using the explanation utilities.

The main examples in this vignette use `SingleCellExperiment`, following Bioconductor conventions. Seurat interoperability is demonstrated separately near the end.

# Input Data

The example below creates a small simulated `SingleCellExperiment` with predicted labels already assigned. In a real analysis, these labels would typically come from an external annotation method such as `SingleR`, Azimuth, or a marker-based annotation workflow.

```{r simulate-data, eval=run_demo}
set.seed(42)

n_genes <- 50
n_cells <- 40

counts <- matrix(
    rpois(n_genes * n_cells, lambda = 5),
    nrow = n_genes,
    ncol = n_cells,
    dimnames = list(
        paste0("Gene", seq_len(n_genes)),
        paste0("Cell", seq_len(n_cells))
    )
)

sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(counts = counts)
)

SummarizedExperiment::colData(sce)$predicted_label <- rep(
    c("T cell", "B cell"),
    each = n_cells / 2
)

sce
```

The predicted cell labels are stored in `colData()`.

```{r sce-metadata, eval=run_demo}
head(SummarizedExperiment::colData(sce))
```

A marker gene database is supplied as a named list, with one element per predicted cell type.

```{r marker-list}
markers <- list(
    "T cell" = c("Gene1", "Gene2", "Gene3"),
    "B cell" = c("Gene4", "Gene5", "Gene6")
)

markers
```

In the simulated object, `predicted_label` already matches these names. `match_labels()`, demonstrated later, can help match more loosely formatted labels.

# Neighborhood Representation

`neighbor_score()` evaluates agreement among nearby cells. For a `SingleCellExperiment`, the object should contain a suitable low-dimensional representation.

For this small example, we create a PCA representation directly from the log-transformed count matrix. This avoids introducing an unrelated preprocessing workflow into the example.

```{r sce-pca, eval=run_demo}
log_counts <- log1p(
    SummarizedExperiment::assay(sce, "counts")
)

pca <- stats::prcomp(
    t(log_counts),
    center = TRUE,
    scale. = TRUE
)$x[, seq_len(5), drop = FALSE]

SingleCellExperiment::reducedDim(
    sce,
    "PCA"
) <- pca

dim(
    SingleCellExperiment::reducedDim(sce, "PCA")
)
```

# Marker Consistency Score

Marker consistency evaluates whether a cell expresses genes expected for its predicted identity. scCertify uses the UCell algorithm to compute rank-based enrichment scores.

```{r marker-score, eval=run_demo}
markers <- list(
    "T cell" = c("Gene1", "Gene2"),
    "B cell" = c("Gene3", "Gene4")
)

SummarizedExperiment::assay(
    sce,
    "logcounts"
) <- log2(
    SummarizedExperiment::assay(
        sce,
        "counts"
    ) + 1
)

marker_scores <- scCertify::marker_score(
    sce,
    markers
)

head(marker_scores)
```

Higher scores indicate stronger molecular support for the assigned annotation.

# Neighborhood Agreement Score

Correctly annotated cells generally occur near cells of the same identity in low-dimensional space.

```{r neighbor-score, eval=run_demo}
neighbor_scores <- neighbor_score(
    sce
)

head(neighbor_scores)
```

Neighborhood agreement ranges from 0 to 1, where larger values indicate greater agreement with nearby cells.

# Entropy-Based Uncertainty

Entropy provides a quantitative measure of annotation uncertainty. Unlike `marker_score()` and `neighbor_score()`, `entropy_score()` operates on a numeric score matrix, such as per-cell prediction probabilities across candidate labels from an external annotation method.

```{r entropy-score, eval=have_entropy}
score_matrix <- matrix(
    runif(20),
    nrow = 5
)

entropy_score(score_matrix)
```

For the simulated dataset, we treat each cell as having a two-class prediction-probability row and normalize the resulting entropy to `[0, 1]`. The normalized value is stored in `colData()` as `entropy_norm`, which is used by `cell_certify()`.

```{r entropy-norm, eval=run_demo}
prob_matrix <- matrix(
    runif(n_cells * 2, min = 0.1, max = 1),
    ncol = 2
)

raw_entropy <- entropy_score(prob_matrix)

SummarizedExperiment::colData(sce)$entropy_norm <-
    raw_entropy / log(ncol(prob_matrix))

head(
    SummarizedExperiment::colData(sce)$entropy_norm
)
```

Low entropy indicates confident predictions, whereas high entropy suggests ambiguous assignments.

scCertify does not currently ship its own doublet detector. A `doublet_score` column can be supplied from an external tool such as `scDblFinder`. Here we simulate one for demonstration.

```{r doublet-score, eval=run_demo}
SummarizedExperiment::colData(sce)$doublet_score <-
    runif(n_cells, min = 0, max = 0.3)
```

If `entropy_norm` or `doublet_score` is missing entirely, `cell_certify()` substitutes zeros and issues a warning rather than failing.

# Computing Confidence Scores

The final confidence score integrates the available evidence through `cell_certify()`. The function internally computes marker and neighborhood scores, combines them with `entropy_norm` and `doublet_score`, and stores the results in `colData()`.

The default weighting scheme is:

| Component | Weight |
|:----------|------:|
| Marker enrichment | 35% |
| Neighborhood agreement | 35% |
| Entropy | 20% |
| Doublet penalty | 10% |

```{r cell-certify, eval=run_demo}
sce <- cell_certify(
    sce,
    markers
)

head(
    SummarizedExperiment::colData(sce)[
        ,
        c(
            "predicted_label",
            "marker_score",
            "neighbor_score",
            "entropy_norm",
            "doublet_score",
            "confidence_score",
            "confidence_class"
        )
    ]
)
```

The resulting `colData()` contains:

- `marker_score`
- `neighbor_score`
- `confidence_score`
- `confidence_class`

`classify_confidence()` can also be used independently with any numeric vector of scores.

```{r confidence-classification}
scores <- c(0.15, 0.55, 0.92)

data.frame(
    score = scores,
    class = classify_confidence(scores)
)
```

# Custom Confidence Thresholds

Confidence classes can be customized by changing the moderate and high thresholds passed to `cell_certify()`.

```{r custom-thresholds, eval=run_demo}
sce_custom <- cell_certify(
    sce,
    markers,
    moderate_threshold = 0.4,
    high_threshold = 0.75
)

table(
    SummarizedExperiment::colData(sce_custom)$confidence_class
)
```

# Explaining Confidence

Two small helper functions support the scoring pipeline and can also be used independently.

```{r helper-functions}
calibrate_confidence(
    c(-1, 0, 1)
)

match_labels(
    "T_cells",
    c("T_cell", "B_cell")
)
```

`calibrate_confidence()` converts raw scores into percentile ranks, while `match_labels()` matches loosely formatted predicted labels against the names of a marker database.

To understand the major contributors to a single cell's confidence score, use `explain_confidence()`.

```{r explain-confidence, eval=run_demo}
explain_confidence(
    sce,
    cell_id = "Cell1"
)
```

For a fuller summary of a single cell, including its predicted label and component scores, use `explain_cell()`.

```{r explain-cell, eval=run_demo}
explain_cell(
    sce,
    cell_id = "Cell1"
)
```

Possible explanations include:

- Weak marker enrichment
- Neighborhood disagreement
- High entropy
- Possible doublet contamination

# Seurat Interoperability

scCertify also supports Seurat objects. The primary workflow above uses `SingleCellExperiment` so that the examples follow Bioconductor conventions; Seurat can be used when it is already part of an analysis workflow.

The following short example demonstrates the same interface without making Seurat preprocessing part of the main workflow.

```{r seurat-interoperability, eval=FALSE}
if (requireNamespace("Seurat", quietly = TRUE)) {

    seurat_object <- Seurat::CreateSeuratObject(
        counts = counts
    )

    seurat_object$predicted_label <- rep(
        c("T cell", "B cell"),
        each = n_cells / 2
    )

    seurat_object$entropy_norm <- runif(n_cells)
    seurat_object$doublet_score <- runif(n_cells)

    seurat_object <- cell_certify(
        seurat_object,
        markers
    )

    head(seurat_object$confidence_score)
}
```

Thus, existing Seurat workflows can use scCertify while Bioconductor users can work directly with `SingleCellExperiment` objects.

# Interpreting Results

High-confidence annotations generally exhibit:

- Strong marker enrichment
- Strong neighborhood agreement
- Low entropy
- Low doublet scores

Low-confidence annotations frequently correspond to:

- Ambiguous transcriptional states
- Rare cell populations
- Doublets
- Poor-quality cells
- Incorrect annotations

These cells may benefit from manual inspection before downstream analyses.

# Summary

scCertify provides an explainable confidence scoring framework for single-cell annotation. By combining marker enrichment, neighborhood agreement, entropy, and doublet information, it quantifies annotation reliability while providing interpretable explanations for individual confidence assignments.

The package is intended to complement existing annotation pipelines and assist with quality control, annotation refinement, downstream filtering, and biological interpretation.

# References

- Andreatta M, Carmona SJ (2021). *UCell: Robust and scalable single-cell gene signature scoring*. Computational and Structural Biotechnology Journal.

- Aran D, Looney AP, Liu L, et al. (2019). *Reference-based analysis of lung single-cell sequencing reveals a transitional profibrotic macrophage*. Nature Immunology.

# Session Information

```{r session-info}
sessionInfo()
```
